Performance Indicators for Shape and Position Assessment in Electromagnetic Inverse Scattering

### 2025-08-12 coverletter/cover_letter.tex
Fixes proposed by Ricardo

\hyphenation{
elec-tro-mag-net-ic 
Trans-ac-tions 
Sam-pling 
geo-met-ric 
con-sid-er-a-tion 
eval-u-a-tion 
re-con-struc-tion 
lo-cal-i-za-tion
algo-rithm
algo-rithms
per-for-mance
indi-ca-tors
scat-ter-ing
micro-wave
bio-med-i-cal
EISPY
in-tro-ducing
con-trast
ap-prox-i-ma-tion
fa-cil-i-tat-ing
}

@@

% Main content
I am writing to submit our manuscript entitled \textbf{``Novel Performance Indicators for Shape and Position Assessment in Electromagnetic Inverse Scattering''} for consideration for publication in IEEE Transactions on Antennas and Propagation.

>>>>

\hyphenation{
elec-tro-mag-net-ic 
Trans-ac-tions 
Sam-pling 
geo-met-ric 
con-sid-er-a-tion 
eval-u-a-tion 
re-con-struc-tion 
lo-cal-i-za-tion
algo-rithm
algo-rithms
per-for-mance
indi-ca-tors
scat-ter-ing
micro-wave
bio-med-i-cal
EISPY
in-tro-ducing
con-trast
ap-prox-i-ma-tion
fa-cil-i-tat-ing
Elec-tro-mag-net-ic
pub-li-ca-tion
con-sid-er-a-tion
As-sess-ment
}

@@

% Main content
I am writing to submit our manuscript entitled \textbf{``Performance Indicators for Shape and Position Assessment in Electromagnetic Inverse Scattering''} for consideration for publication in IEEE Transactions on Antennas and Propagation.

### 2025-08-12 paper/main.tex
Fixes proposed by Ricardo

    \title{Novel Performance Indicators for Shape and Position Assessment in Electromagnetic Inverse Scattering}
    \author{Andr\'e Costa Batista, Ricardo Adriano, and Lucas S. Batista
    \thanks{This work was supported in part by the Brazilian agency CAPES (Coordination for the Improvement of Higher Education Personnel) under Grant 88887.463864/2019-00, the scholarship FAPEMIG-CNPQ (process APQ-06716-24), and CNPq (The National Council for Scientific and Technological Development).}
    \thanks{A. C. Batista, R. Adriano, and L. S. Batista are with the Department of Electrical Engineering, Universidade Federal de Minas Gerais, Belo Horizonte, MG 31270-901, Brazil (e-mail: andre-costa@ufmg.br; rluiz@ufmg.br; lusoba@ufmg.br). A. C. Batista and L. S. Batista are also with Operations Research and Complex Systems Laboratory (ORCS Lab), Belo Horizonte, MG, Brazil.}}

@@

    \begin{abstract}
        Microwave imaging is a low-cost, non-invasive technique for detecting and characterizing objects in inaccessible media through electromagnetic field measurements. The evaluation of algorithms that solve the related electromagnetic inverse scattering problem typically relies on metrics such as Mean Square Error (MSE) or Structural Similarity (SSIM), which primarily assess contrast estimation accuracy but provide limited insight into the recovery of object geometry and position. This paper introduces two novel performance indicators specifically designed to evaluate electromagnetic inverse scattering algorithms: a shape error indicator that quantifies geometric reconstruction accuracy, and a position error indicator that measures object localization precision. The proposed indicators are applicable to both qualitative and quantitative methods, even when they are combined in an experiment. Comprehensive experimental validation is conducted using traditional algorithms: Linear Sampling Method (LSM), Orthogonality Sampling Method (OSM), Born Iterative Method (BIM), Contrast Source Iterative Method (CSI), Subspace Optimization Method (SOM), and Circle Approximation (CA). Three distinct experimental frameworks are employed: shape recovery studies, position detection studies, and a realistic breast phantom study demonstrating their applicability. Supported by robust statistical analysis, the experiments revealed that: (a) in scenarios with a single scatterer and low Degree of Nonlinearity, BIM achieves the best average performance in object geometry recovery; and (b) in scenarios with a single small-sized, high-contrast scatterer, SOM and OSM achieve the best average performance in target localization. These indicators provide useful tools for algorithm development, benchmarking studies, and performance assessment in electromagnetic inverse scattering applications.
    \end{abstract}

@@

        It is also worth highlighting the Structural Similarity (SSIM) metric \cite{wang2004image} to quantitatively show the reconstruction visual quality, which is designed to compute the mismatch between reconstructed contrast results and corresponding ground truths both on the pixel-wise level and perceptual level \cite{wang2023push,liu2025exploring}. Instead of performing a pixel-by-pixel comparison, this metric considers aspects of luminance, contrast, and structure of any given image. Therefore, while it provides a deeper quantification of overall reconstruction quality, it does not directly measure isolated aspects such as geometry and position.

@@

            This approach has advantages and disadvantages. If the reconstructed image contains multiple scatterers, and one or more of them has a contrast below this threshold, the methodology will fail to detect the contour of these objects. However, in scenarios where there is only one scatterer in the image, this approach is simple and effective. In fact, a single scatterer is the most objective way to test an algorithm's potential in recovering geometry, since the presence of multiple scatterers can degrade the algorithm's performance due to the influence of the induced currents in each scatterer on the scattering of the others. Furthermore, this approach poses no issues for algorithms capable of accurately capturing contours, such as machine learning-based methods. Moreover, instead of using the values from the original image, it is more beneficial to use only the values from the reconstructed image, both to isolate the influence of error in the contrast estimate and to apply the metric to qualitative methods that do not estimate contrast. If, for some reason, the original contrast function involves soft boundaries of the objects, the threshold might also be applied to the original image.

@@

            Considering this, the experiments in this section were designed for scenarios involving strong scatterers with small size, where traditional methods cannot adequately recover scatterer geometry. Therefore, the position error indicator $\zeta_P$ serves as a relevant metric for comparing algorithm performance in such challenging scenarios. In addition, the DOI discretization for data synthesis was performed using a finer grid with $160 \times 160$ pixels.

            Considering this, the experiments in this section were designed for scenarios involving strong scatterers with small dimensions, where traditional methods cannot adequately recover scatterer geometry. Therefore, the position error indicator $\zeta_P$ serves as a relevant metric for comparing algorithm performance in such challenging scenarios.

            For all subsequent experiments, LSM was removed and an other method was introduced. Since the analytical solution for scattering from an infinite dielectric cylinder is well-established \cite{harrington2001time}, it becomes possible to determine the optimal circular approximation (defined by radius, position, and contrast) that best fits the scattered field of any single scatterer.

>>>>

    \title{Performance Indicators for Shape and Position Assessment in Electromagnetic Inverse Scattering}
    \author{Andr\'e Costa Batista, Ricardo Adriano, and Lucas S. Batista
    \thanks{This work was supported in part by the Brazilian agency CAPES (Coordination for the Improvement of Higher Education Personnel) under Grant 88887.463864/2019-00, the scholarship FAPEMIG-CNPQ (process APQ-06716-24), and CNPq (The National Council for Scientific and Technological Development).}
    \thanks{A. C. Batista, R. Adriano, and L. S. Batista are with the Department of Electrical Engineering, Universidade Federal de Minas Gerais, Belo Horizonte, MG 31270-901, Brazil (e-mail: andre-costa@ufmg.br; rluiz@ufmg.br; lusoba@ufmg.br). A. C. Batista and L. S. Batista are also with Operations Research and Complex Systems Laboratory (ORCS Lab), Belo Horizonte, MG, Brazil.}}

@@

    \begin{abstract}
        Microwave imaging is a low-cost, non-invasive technique for detecting and characterizing objects in inaccessible media through electromagnetic field measurements. The evaluation of algorithms that solve the related electromagnetic inverse scattering problem typically relies on metrics such as Mean Square Error (MSE) or Structural Similarity (SSIM), which primarily assess contrast estimation accuracy but provide limited insight into the recovery of object geometry and position. To address this limitation, this paper introduces two novel performance indicators specifically designed to evaluate electromagnetic inverse scattering algorithms: a shape error indicator that quantifies geometric reconstruction accuracy, and a position error indicator that measures object localization precision. The proposed indicators are applicable to both qualitative and quantitative methods, even when they are combined in an experiment. Comprehensive experimental validation is conducted using traditional algorithms: Linear Sampling Method (LSM), Orthogonality Sampling Method (OSM), Born Iterative Method (BIM), Contrast Source Iterative Method (CSI), Subspace Optimization Method (SOM), and Circle Approximation (CA). Three distinct experimental frameworks are employed: shape recovery studies, position detection studies, and a realistic breast phantom study demonstrating their applicability. Supported by robust statistical analysis, the experiments revealed that: (a) in scenarios with a single scatterer and low Degree of Nonlinearity, BIM achieves the best average performance in object geometry recovery; and (b) in scenarios with a single small-sized, high-contrast scatterer, SOM and OSM achieve the best average performance in target localization. These indicators provide useful tools for algorithm development, benchmarking studies, and performance assessment in electromagnetic inverse scattering applications.

    \end{abstract}

@@

        It is also worth highlighting the Structural Similarity (SSIM) metric \cite{wang2004image} for quantifying reconstruction visual quality, which is designed to compute the mismatch between reconstructed contrast results and corresponding ground truths, both at the pixel-wise level and perceptual level \cite{wang2023push,liu2025exploring}. Instead of performing a pixel-by-pixel comparison, this metric considers aspects of luminance, contrast, and structure of any given image. Therefore, while it provides a deeper quantification of overall reconstruction quality, it does not directly measure isolated aspects such as geometry and position.

@@

            This approach has advantages and disadvantages. If the reconstructed image contains multiple scatterers where one or more has a contrast below this threshold, the methodology will fail to detect the contour of these objects. However, in scenarios with a single scatterer, this approach is simple and effective. Single scatterer scenarios represent the most objective way to test an algorithm's geometry recovery potential, since multiple scatterers can degrade performance due to mutual current induction effects between objects.

            This approach poses no issues for algorithms capable of accurately capturing contours, such as machine learning-based methods. Furthermore, using only values from the reconstructed image (rather than the original image) helps isolate contrast estimation errors and enables metric application to qualitative methods that do not estimate contrast. When the original contrast function involves soft object boundaries, the threshold may also be applied to the original image for consistency.

@@

            Considering this, the experiments in this section were designed for scenarios involving strong scatterers with small dimensions, where traditional methods cannot adequately recover scatterer geometry. Therefore, the position error indicator $\zeta_P$ serves as a relevant metric for comparing algorithm performance in such challenging scenarios. In addition, the DOI discretization for data synthesis was performed using a finer grid with $160 \times 160$ pixels.

            For all subsequent experiments, LSM was removed and an other method was introduced. Since the analytical solution for scattering from an infinite dielectric cylinder is well-established \cite{harrington2001time}, it becomes possible to determine the optimal circular approximation (defined by radius, $x$-$y$ center position, and contrast) that best fits the scattered field of any single scatterer.
